Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
changes/edits/comments in sections 4 and 9

What about actual software engineering principles, such as modularity and
extensibility? This is how industry maintains software, and ensures it continues
to be useful. Everyone \note{everyone in the audience?  paper authors?  software developers in general?} thinks that you can't rewrite your software, but if it is
modular then you can keep systems up to date. Extensibility will keep you
relevant if it is built into the project. One counterpoint raised by Jason Riedy\todo{add cite}
was that trying to take advantage of the latest and greatest hardware often
makes this painful, hence the lack of papers mentioning ``GPUs and exotic
hardware''.

The question of whether funders, such as the NSF, can mandate software plans in
much the same way as they do data management plans. Daniel Katz responded that
software is supposed to be described as part of the NSF data management plan,
and that in their definition data includes software. A comment from Twitter
(@biomickwatson) raised the issue that this requires reviewers and funders who
understand the answers that are given in these plans. Daniel Katz responded
that in programs focused on software or data this can be done effectively, but
conceded that in more general programs this is indeed a problem.

@@

Some of the questions raised after the panel discussion.

@@

\subsection{Papers}

@@

\section{Cross-cutting Issue: Defining Sustainability \scriptsize{\note{lead: Daniel S. Katz}}}  \label{sec:cross-cutting}

The question of what was meant by ``sustainability'' in the context of software
came up in many different parts of the workshop, specifically in the first
keynote (Section~\ref{sec:keynote1}), the Developing and Maintaining Software panel, and the Policy panel.

@@

For the purpose of studying software case studies, we classify the software
projects discussed in the workshop in three broad categories. First, 
`utility software', software tools that enable and/or facilitate the
development of other tools and techniques to carry out scientific work. Second,
`scientific software', software that was developed with an aim to solve a
class of scientific problems. Third, `infrastructure software', software
developed to efficiently utilize new research infrastructure.

These categories have some overlap, and in some \note{most?} cases, the
case studies do not cleanly fit into a single category.
Our classification is based on two criteria: first, the character
of a software project that dominates and best fits, and, second,
the theme of the workshop discussion in which the paper discussing
the project was placed.

@@

\subsection{Scientific Software}
Special purpose software that was developed specifically for a target use-case
scenario or keeping some fixed/frozen requirements in mind. Software projects
pertaining to specific scientific domain often tend to be in a niche and the
user community tends to be small to medium. Numerical accuracy and algorithmic
optimization are some of the paramount requirements of most scientific
software. Advantages of a smaller codebase and requirements include stability,
ease of installation, configurability and integration into larger systems.
While the software can stay stable and require relatively low maintenance, the
responsibility is often on the shoulders of a very few developers. Community
participation in the development tends to be low. Development is linear and
simplistic with a limited scope to follow software best practices. Domain
science oriented software developed to solve  specific problems, e.g.,
\emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. Sustainability
of such software is often a significant challenge. Many submissions report the
software considered as a `byproduct' of actual research.  Others contend that
the software was not the main funded part of their research. One exception is
\emph{Kitware}. While being a software product specializing in scientific
process, developing communities around software process is a core development
focus. One example of this process is the development of the \emph{CMake} build
utility which started out as a building tool for \emph{ITK} but grown to become
a generic build utility for C++ projects.

\subsection{Infrastructure Software}
%TODO: expand
Software developed to better utilize new infrastructure and architectures,
e.g., \emph{MVAPICH}, \emph{visIT}. Explores uncharted territories. Highest
risk since the projects are experimental. Successful projects reap high rewards
and have longer usage span. By nature, they overlap with the utility software
as they must serve as a facilitator to run ordinary codes. Long
requirement-development-test cycles.  Wide community support but tends to be
handled by specialists rather then end users. Promotes collaborations across
the breadth (eg. different departments) and depth (eg. stakeholders within a
department) of community, one of the key ingredient of a sustainable process.
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
associated with software projects dealing with new architectures. To leverage
the power of accelerators such as GPUs and MICs new code and libraries are
required. The experience of efforts as described in~\cite{Ferenbaugh_WSSSPE}
met with a limited success but nonetheless with many invaluable lessons were
learned about influential cultural and technical aspects in sustainable software
development practices.

>>>>

What about actual software engineering principles, such as modularity and
extensibility? This is how industry maintains software, and ensures it continues
to be useful. Everyone \note{everyone in the audience?  paper authors?  software developers in general?} thinks that you can't rewrite your software, but if it is
modular then you can keep systems up to date. Extensibility will keep you
relevant if it is built into the project. One counterpoint raised by Jason Riedy
was that trying to take advantage of the latest and greatest hardware often
makes this painful, hence the lack of papers mentioning ``GPUs and exotic
hardware''.

The question of whether funders, such as the NSF, can mandate software plans in
much the same way as they do data management plans. Daniel Katz responded that
software is supposed to be described as part of the NSF data management plan,
and that in their definition, data includes software. A comment from Twitter
(@biomickwatson) raised the issue that this requires reviewers and funders who
understand the answers that are given in these plans. Daniel Katz responded
that in programs focused on software or data this can be done effectively, but
agreed that in more general programs this is indeed a problem.

@@

Some of the questions raised after the panel discussion are:

@@

\subsection*{Papers}

@@

\section{Cross-cutting Issue: Defining Sustainability \note{lead: Daniel S. Katz}}  \label{sec:cross-cutting}

The question of what was meant by ``sustainability'' in the context of software
came up in many different parts of the workshop, specifically in the first
keynote (\S\ref{sec:keynote1}), the Developing and Maintaining Software panel, and the Policy panel.

@@

For the purpose of studying software case studies, we classify the software
projects discussed in the workshop in three broad categories. The first is 
`utility software', software tools that enable and/or facilitate the
development of other tools and techniques to carry out scientific work. The second is
`scientific software', software that was developed with an aim to solve a
class of scientific problems. The third is `infrastructure software', software
developed to efficiently utilize new research infrastructure.

These categories have some overlap, and in some cases, the
case studies do not cleanly fit into a single category.
Our classification is based on two criteria: first, the character
of a software project that dominates and where it `seems' to be the best fit, and, second,
the the workshop discussion session where the paper about
the project was discussed.

@@

\subsection{Scientific software}
Scientific software consists primarily of special purpose software that was developed for a target use-case
scenario or keeping some fixed/frozen requirements in mind. Software projects
pertaining to specific scientific domain often tend to be in a niche and the
user community tends to be small to medium. Numerical accuracy and algorithmic
optimization are some of the paramount requirements of most scientific
software. Advantages of a smaller codebase and requirements include stability,
ease of installation, configurability and integration into larger systems.
While the software can stay stable and require relatively low maintenance, the
responsibility is often on the shoulders of a very few developers. Community
participation in the development tends to be low. Development is linear and
simplistic with a limited scope to follow software best practices. Domain
science oriented software developed to solve  specific problems, e.g.,
\emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. Sustainability
of such software is often a significant challenge. Many submissions reported the
software considered a `byproduct' of the actual research.  Others contended that
the software was not the main funded part of their research. One exception was
\emph{Kitware}, which while being a software product specializing in scientific
process, has a core focus of developing communities around software processes.
One example of this process is the development of the \emph{CMake} build
utility, which started out as a building tool for \emph{ITK} but grew to become
a generic build utility for C++ projects.

\subsection{Infrastructure software}
%TODO: expand
Infrastructure software is developed to better utilize particular, often new, infrastructures and architectures,
e.g., \emph{MVAPICH}, \emph{visIT}. Explores uncharted territories. Highest
risk since the projects are experimental. \note{Dan: so is the only difference here that
this category is aimed at new systems, where utility software is not?  If so, perhaps this should be incorporated into that subsection, so there would just be two subsections. Something in the framing of this subsection doesn't feel right to me.} Successful projects reap high rewards
and have longer usage span. By nature, they overlap with the utility software
as they must serve as a facilitator to run ordinary codes. Long
requirement-development-test cycles.  Wide community support but tends to be
handled by specialists rather then end users. Promotes collaborations across
the breadth (e.g., different departments) and depth (e.g., stakeholders within a
department) of community, one of the key ingredient of a sustainable process.
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
associated with software projects dealing with new architectures. To leverage
the power of accelerators such as GPUs and MICs new code and libraries are
required. The experience of efforts as described in~\cite{Ferenbaugh_WSSSPE}
met with a limited success but nonetheless with many invaluable lessons were
learned about influential cultural and technical aspects in sustainable software
development practices.
